Bulk Semantic Annotation with a Partially Known Knowledge Base

### 2024-12-11 doc/cicm25-sysdesc/sysdesc.tex
more

\paragraph{Traditional, informal authoring}
In traditional authoring e.g. of a scientific article or a textbook, the domain model -- a
highly structured model of the domain of discourse -- is in the authors' heads, and is
partially verbalized e.g. in the preliminaries section or the main sections of the
respective article. The content is informal -- usually a natural language like english
augmented with technical jargon, formulae, tables, and diagrams -- and designed for
processing via the brain of a colleague or student -- in any case, a human who shares a
canonical part of the domain model with the authors.\ednote{MK: this shared domain model
  could be thought of as the informal content commons; hmmm, but maybe the literature is
  the content commons} Consequently a large part of the authoring problem lies in
predicting the state and extent of the domain model of the reader and creating text that
bridges the gap between the reader's domain model and the payload content of the article
-- the knowledge the article or textbook intends to impart -- ideally using the technical
vocabulary the reader is familiar with to reduce this gap. Tool support for traditional
authoring is minimal and usually restricted to spell/grammar-checkers and online thesauri
if we disregard LLM writing/formulation support for now.

\paragraph{Formal Authoring}
In fully formal authoring e.g. for programming, the content commons is given formally in
the form of the libraries of the respective programming language, correspondingly the
authored content is also formal -- a program -- and intended for processing by a computer,
for which the formality is a prerequisite. Formal authoring is usually supported by an
integraded development environment (IDE) that makes use of the fully formal content and
offers services like semantic highlighting, tab completion, jumping to variable
declarations, function definitions, and interface definitions, as well as refactoring
support. All of these rely on a syntactic/semantic analysis of the formal content. The
situation is very similar for authoring of formal knowledge, e.g. in program verification
of the formalization of mathematics.

@@

But in addition to the base text, semantic authoring also has to provide the semantic
annotations that drive the personalization and interaction. Techniques from formal
authoring could help here, but \ednote{cite Luka's CICM paper \cite{VreWelKam:tsmmdui24}
  for the formulae; introduce the \sTeX way of writing content formulae above.} But in
contrast to formal authoring, we cannot just trigger the IDE functionality by the formal
syntax, as the authored content is -- so far -- informal and the content commons (the
target of the annotations) is itself only flexiformal. In a way, we need a local
flexiformalization workflow that helps authors bridge the informal-formal gap.

\paragraph{Semantic Authoring in \sTeX}
How this can concretely work depends on the particular semantic framework, the specific
annotations we want to support, and the semantics embedded in the content commons.

@@

One practical problem with \snify that remains is that the system only harvests
symbol/verbalization pairs from \sTeX files on the local file system. This means that for
effective annotation users (have to) download the whole content commons and keep it
updated. Unfortunately, the current MMT-based MathHub knowledge management
API\ednote{intro this above} does not supply verbalization information. The iMMT-based
API~\cite{iMMT:on} currently under development will, so \snify will be able to request the
pairs for the remaining (non-local) content commons. If a pair is chosen in an annotation,
\snify can then automatically download the corresponding archive so that the annotated
file can be formattted with pdflatex or rusTeX locally.

@@

In this paper we have presented the semantic authoring problem using the \sTeX corpus and
the adaptive learning assistant \ALeA as a concrete example. We have shown that it differs
from both classical authoring and formal authoring in terms of the necesssary system
support. As an example of specialized authoring support, we have presented \snify, a
command-line tool that bridges the gap between formal symbols in \sTeX and

@@

% LocalWords:  tbw Aarne Ranta homostem pre

>>>>

\paragraph{Traditional, informal authoring}
In traditional authoring e.g. of a scientific article or a textbook, the domain model -- a
highly structured model of the domain of discourse -- is in the authors' heads, and is
partially verbalized e.g. in the preliminaries section or the main sections of the
respective article. The content is informal -- usually a natural language like English
augmented with technical jargon, formulae, tables, and diagrams -- and designed for
processing via the brain of a colleague or student -- in any case, a human who shares a
canonical part of the domain model with the authors.\ednote{MK: this shared domain model
  could be thought of as the informal content commons; hmmm, but maybe the literature is
  the content commons} Consequently a large part of the authoring problem lies in
predicting the state and extent of the domain model of the reader and creating text that
bridges the gap between the reader's domain model and the payload content of the article
-- the knowledge the article or textbook intends to impart -- ideally using the technical
vocabulary the reader is familiar with to reduce this gap. Tool support for traditional
authoring is minimal and usually restricted to spell/grammar-checkers and online thesauri
if we disregard LLM-based writing/formulation support for now.

\paragraph{Formal Authoring}
In fully formal authoring e.g. for programming, the content commons is given formally in
the form of the libraries of the respective programming language, correspondingly the
authored content is also formal -- a program -- and intended for processing by a computer,
for which the formality is a prerequisite. Formal authoring is usually supported by an
integrated development environment (IDE) that makes use of the fully formal content and
offers services like semantic highlighting, tab completion, jumping to variable
declarations, function definitions, and interface definitions, as well as refactoring
support. All of these rely on a syntactic/semantic analysis of the formal content. The
situation is very similar for authoring of formal knowledge, e.g. in program verification
of the formalization of mathematics.

@@

But in addition to the base text, semantic authoring also has to provide the semantic
annotations that drive the personalization and interaction. Techniques from formal
authoring could help here, but in contrast to formal authoring, we cannot just trigger the
IDE functionality by the formal syntax, as the authored content is -- so far -- informal
and the content commons (the target of the annotations) is itself only flexiformal. In a
way, we need a local flexiformalization workflow that helps authors bridge the
informal-formal gap.

\paragraph{Semantic Authoring in \sTeX}
How this can concretely work depends on the particular semantic framework, the specific
annotations we want to support, and the semantics embedded in the content commons. In the
\sTeX framework, where we semantically annotate formulae and technical terms with pointers
to their definienda, we can bridge the presentation/semantics gap -- we annotate the words
in sentences and glyphs in formulae with the verbalizations and notations in the content
commons.\ednote{MK: maybe explain them again}. A tool for supporting semantic authoring of
formulae via notations has been presented by Vre{\v{c}}ar et al. in
\cite{VreWelKam:tsmmdui24}. In this paper we present a tool that supports semantic
annotation of technical terms via verbalizations.

@@

One practical problem with \snify that remains is that the system only harvests
symbol/verbalization pairs from \sTeX files on the local file system. This means that for
effective annotation users (have to) download the whole content commons and keep it
updated. Unfortunately, the current MMT-based MathHub knowledge management
API\ednote{intro this above} does not supply verbalization information. The iMMT-based
API~\cite{iMMT:on} currently under development will, so \snify will be able to request the
pairs for the remaining (non-local) content commons. If a pair is chosen in an annotation,
\snify can then automatically download the corresponding archive so that the annotated
file can be formatted with pdflatex or rusTeX locally.

@@

In this paper we have presented the semantic authoring problem using the \sTeX corpus and
the adaptive learning assistant \ALeA as a concrete example. We have shown that it differs
from both classical authoring and formal authoring in terms of the necessary system
support. As an example of specialized authoring support, we have presented \snify, a
command-line tool that bridges the gap between formal symbols in \sTeX and

@@

% LocalWords:  tbw Aarne Ranta homostem pre hmmm Vre ar al sn ify
